---
title: 'Evaluate decoy quality'
output: html_notebook
author: Lieven Clement
---

#Data Import
##Load data
mzid files can be loade via de mzID function in the mzID package.
Because we prefer to work with a the data in a table format we flatten the data.
We download the file from the web, but you can also replace the text string by a path to the file on your local computer.
We show the first 6 rows of data.

```{r}
library(mzID)
library(tidyverse)
dataFlat <- mzID("https://raw.githubusercontent.com/statOmics/SGA2019/data/identification/peptideShakerTutorial.mzid") %>% flatten
head(dataFlat)
```

##Column names

We will assess the names of the variables in the mzid file.

```{r}
names(dataFlat)
```

#Evaluate decoys
We would like to assess the decoy quality for the omssa and x!tandem scores.
The scores are e-values.
Small e-values are better.
For visualisation purposes we transform the e-values by taking the -log10

##Omssa
###Preprocess mzid data
We first subset the data, because there are many missing values: PSMs without omssa score, and we remove the missing values.
Next we -log10 transform the data.

```{r}
data <- dataFlat[,c("isdecoy","omssa:evalue")]
names(data) <- c("decoy","score")
data <- na.exclude(data)
data$score <- -log10(as.double(data$score))
```

###Histogram of targets and decoys
```{r}
library(ggplot2)

#We will use 50 breaks
nBreaks <- 50
binwidth <- diff(range(data$score))/nBreaks
print(binwidth)
      ggplot(data, aes(score, fill = decoy, col=I("black")))+ geom_histogram(alpha = 0.5, binwidth=binwidth, position = 'identity') +  labs(x = 'Score', y = 'Counts' ,title = 'Histogram of targets and decoys') +
  theme_bw() +
  theme(
    plot.title = element_text(size = rel(1.5)),
    axis.title = element_text(size = rel(1.2)),
    axis.text = element_text(size = rel(1.2)),
    axis.title.y = element_text(angle = 0))
```

###PPplot of targets and decoys

```{r}
pi0 <- sum(data$decoy)/sum(!data$decoy)
ppPlot <- ggplot()  +
geom_abline(slope = pi0,color = 'black') +
labs(x = 'Decoy Percentile', y = 'Target\nPercentile' ,title = 'PP plot of target PSMs') +
xlim(0,1) + ylim(0,1) +
theme_bw() +
theme(
   plot.title = element_text(size = rel(1.5)),
   axis.title = element_text(size = rel(1.2)),
   axis.text = element_text(size = rel(1.2)),
   axis.title.y = element_text(angle = 0))

x <- data$score[!data$decoy]
Ft <- ecdf(x)
Fd <- ecdf(data$score[data$decoy])
df <- data_frame(Fdp = Fd(x), Ftp = Ft(x))

ppPlot + geom_point(data = df,aes(Fdp,Ftp),color = 'dark grey')
```

 ##X!tandem
 Copy the code above and adopt it so as to evaluate the X!tandem decoys.
